\section{DEFINING SPECIALIZED CYBER RED TEAM RESPONSIVE OPERATIONS}
\label{sec:crtopreq}
\glsresetall
Nations are not well known to announce publicly their offensive cyber capacities and more focused on advertising and promoting cyber defence activities and initiatives. Despite this, the identified major cyber intelligence programmes attributed to nation states, such as, US NSA PRISM\footnote{The Verge. ``Everything you need to know about PRISM.'' \url{https://www.theverge.com/2013/7/17/4517480/nsa-spying-prism-surveillance-cheat-sheet}. Accessed: 28/09/2018} and UK GCHQ TEMPORA\footnote{Wired. ``A simple guide to GCHQ's internet surveillance programme Tempora.'' \url{https://www.wired.co.uk/article/gchq-tempora-101}. Accessed: 28/09/2018}, prove to represent the actual cyber capabilities of some NATO alliance members.
Also, collective defence alliances and treaties as of now do not acknowledge cyber offence as a collective approach and leave such initiatives up to every individual nation to assess and develop on its own. However, this does not exclude any other allies either requesting such assistance or joining forces to engage in offensive cyber operations.
Furthermore, responsive cyber defence balances between the defensive and offensive posture and can enable defenders to engage the adversary both in their own cyber infrastructure, as well as globally. However, there is no clear understanding on how such activities can be executed and what their operational and technical requirements are.

This chapter addresses responsive activities taken in the cyberspace by introducing and defining the concept of \textit{specialized cyber red team responsive computer network operations}. This concept is divided into three main components: 1) responsive cyber defence -- the overarching concept defining the defender's posture to pursue the adversary to ensure the maximum possible protection of the defended assets; 2) computer network operations -- are the means of the defender to exercise this posture against the adversary and systems engaged in the attack; and 3) cyber red team -- the actual team of experts possessing the required skills and capabilities to engage in such responsive operations.

The following subsections present the individual concepts of responsive cyber defence, computer network operations, cyber red teaming, and assesses their minimal operational requirements.
These principles lay the required foundation to define, once combined, the specialized cyber red team responsive operations and specify their requirements.

\input{sections/core/crtopreq/rcd}
\input{sections/core/crtopreq/cno}
\input{sections/core/crtopreq/crt}


\section{SPECIALIZED CYBER RED TEAM RESPONSIVE OPERATIONS}
\label{sec:crtops}
\glsresetall
Concepts, methods and approaches presented in the listed publications (\ref{pub:firstPub}, \ref{pub:secondPub}, and \ref{pub:thirdPub}) are applied to fulfil the requirements of specialized cyber red team responsive cyber operations, allowing the creation of custom cyber red team tools and techniques for the responsive operation execution. Furthermore, such technique applicability is evaluated and proposed to counter the cyber kill-chain. Additionally, an insight into legal ramifications of specialized cyber red team responsive operations is presented.

\input{sections/core/crtops/scrtops}
\input{sections/core/crtops/counterkill}

% Further chapters
\input{sections/core/crtops/protection}
\input{sections/core/crtops/exercises}


\section{CONCLUSIONS AND FUTURE WORK}
\label{sec:end}
\glsresetall
\input{sections/conclusions}